Notes of Basic Topolgy

### 2016-11-28 Basic Topology Notes/sections/ch5-fundamental-groups.tex
[minor] physicist-taper: add some symbols (N,E). [minor] Basic Topology: add the comb space example to homotopy type and stess not necessarily path-connected there. Clearify one todo about straight-line homotopy on S^n

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{pics/homotopy-for-loops-2.PNG}
    \caption{Homotopy for loops 2 (pp. 88 of \cite{book})}
\end{figure}
\begin{ex}
    The author shows on page 88 of \cite{book} that: when $C$ is a
    convex subset of a euclidean space, let $f,g:X\to C$ be continuous
    maps, then $f\simeq_F g$, where $F$ is $F(x,t)=(1-t)f(x)+tg(x)$.
    Note that if $f$ and $g$ agree on a subset $A$ of $X$, then this
    homotopy is a homotopy relative to $A$. This $F$ is called a
    \nomen{straight-line homotopy}.
\end{ex}
\begin{ex}
    Let $f,g:X\to S^n$ be continuous maps. We can take $S^n$ to be the
    unit sphere in $\mathbb{E}^{n+1}$, and think of $f,g$ as
    continuous maps into $\mathbb{E}^{n+1}$, then we may form a
    straight-line homotopy from $f$ to $g$ by:
    \begin{equation}
        F(x,t) = \frac{(1-t)f(x)+tg(x)}{||(1-t)f(x)+tg(x)||}
    \end{equation}
    But why we are  % yes why! TODO
\end{ex}
\begin{ex}
    This is example is best illustrated by pictures:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{pics/{loops-to-loops-fig5.2}.PNG}
        \caption{Loops to loops (pp. 89 of \cite{book})}
    \end{figure}
    Geometrieally, $\alpha$ winds eaeh of the segments
    $[O,\frac{1}{2}]$, $[\frac{1}{2},\frac{3}{4}]$, $[\frac{3}{4}, 1]$
    once round the eirc1e, the first two being wound in an
    anticlockwise direetion, and the third clockwise. The loop $\beta$
    simply winds the whole interval $[0,1]$ once round the circle
    anticlockwise.

@@

\begin{defi}[Homotopy type]
\nomenclature{Homotopy type}{\nomrefpage.}
    Two spaces $X$ and $Y$ have the same homotopy type, or they are
    homotopy equivalent, if there exist maps:
    $$ \begin{tikzcd}[]
        X\ar[r,"f",bend left] & Y\ar[l,"g",bend left]
    \end{tikzcd}$$
    such that $g\circ f\cong \id_X$, and $f\circ g\cong\id_Y$.
\end{defi}
\begin{fact}
This relationship must be an equivalence relation on topological
spaces, as confirmed by lemma (5.16) on page 103 of \cite{book}.
\end{fact}

\begin{defi}[Deformation retraction]
\nomenclature{Deformation retraction}{\nomrefpage.}
    Let $A$ be a subspace of $X$. Let a homotopy $G:X\times I\to X$ which
    is relative to $A$ and for which
    $$ \begin{cases}
        G(x,0)=x & \\
        G(x,1)\in A
    \end{cases}$$
    for all $x\in X$. Then $G$ will be called a deformation retraction
    of $X$ onto $A$.
\end{defi}
\begin{remark}
    If there is a deformation retraction of $X$ onto $A$, then of
    course $X$ and $A$ have the same homotopy type.
\end{remark}
Following is an example of a deformation retraction:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{pics/deformation-retraction.PNG}
    \caption{Three Deformation Retractions (from \cite{book})}
\end{figure}

@@

\begin{thm}
    If two path-connected spaces are of the same homotopy type, then
    they have isomorphic fundamental group.
\end{thm}
\begin{proof}
    Proof can be found on page 106 of \cite{book}. Here's a note using
    the notation in that book:
    $$ \begin{tikzcd}[]
        X\ar[r,"f",bend left] & Y\ar[l,"g",bend left]
    \end{tikzcd}$$
    It should be noticed that the continuity allows one to identify the
    diagram:
    $$ \begin{tikzcd}[]
        \pi_1(X,p)\ar[r,"(gf)_*"]\ar[rr,"\id",bend right]
            &\pi_1(X,gf(p))\ar[r,"\gamma^{-1}_*"]
            &\pi_1(X,p)
    \end{tikzcd} $$
    with the diagram:
    $$ \begin{tikzcd}[]
        \pi_1(X)\ar[r,"(gf)_*"]\ar[rr,"\id",bend right]
            &\pi_1(X)\ar[r,"\gamma^{-1}_*"]
            &\pi_1(X)
    \end{tikzcd} $$
\end{proof}

\begin{fact}
Using above fact, we can find that the M\"obius strip, the cylinder,
the punctured plane $\E^2\setminus\{0\}$, and the solid torus, all
have the homotopy type of a circle $S^1$, and consequiently have $\Z$
as fundamental group.
\end{fact}
\begin{fact}
    Also, $\E^n\setminus\{0\}$ deformation-retracts onto $S^{n-1}$ and
    is therefore a simply connected space when $n\geq 3$.
\end{fact}

@@

    \begin{enumerate}
        \item A space is contractible if and only if it has the
            homotopy type of a point.
        \item A contractible space is simply connected.
        \item Any two maps into a contractible space are homotopic.
        \item If $X$ is contractible, then $\id_X$ is homotopic to the
            constant map at $x$ for any $x\in X$.
    \end{enumerate}
\end{thm}
\begin{ex}
    Here is a contractible space that is really hard to imagin. It is
    called the topologist's dunce hat. It is formed by identifying the
    sides of a triangle in the manner indicated in the following
    figure:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.4\linewidth]{pics/topo-dunce-hat.PNG}
        \caption{Topologist's dunce hat}
        \label{fig:topo-d-hat}
    \end{figure}
    You may imagin identify a pair of sides, then identify the
    identitied edge to the resting side. The direction for identifying
    is really important.
    This space is contractible. Despite some hard words said by the
    book, it is really simple to see on figure~\ref{fig:topo-d-hat}.
    Draw any loop on it and shrink it. Noticing that all edges are
    identified, though in some odd direction.
\end{ex}

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{pics/homotopy-for-loops-2.PNG}
    \caption{Homotopy for loops 2 (pp. 88 of \cite{book})}
\end{figure}
\begin{ex}
    The author shows on page 88 of \cite{book} that: when $C$ is a
    convex subset of a euclidean space, let $f,g:X\to C$ be continuous
    maps, then $f\simeq_F g$, where $F$ is $F(x,t)=(1-t)f(x)+tg(x)$.
    Note that if $f$ and $g$ agree on a subset $A$ of $X$, then this
    homotopy is a homotopy relative to $A$. This $F$ is called a
    \nomen{straight-line homotopy}.
\end{ex}
\begin{ex}
    Let $f,g:X\to S^n$ be continuous maps. We can take $S^n$ to be the
    unit sphere in $\mathbb{E}^{n+1}$, and think of $f,g$ as
    continuous maps into $\mathbb{E}^{n+1}$, then we may form a
    straight-line homotopy from $f$ to $g$ by:
    \begin{equation}
        F(x,t) = \frac{(1-t)f(x)+tg(x)}{||(1-t)f(x)+tg(x)||}
    \end{equation}
    Notice that the ball $B^{n+1}$ is a convex set, so the numerator
    lies inside the ball. When normalized (as in $F(x,t)$), the
    numerator is a point on the sphere.
\end{ex}
\begin{ex}
    This is example is best illustrated by pictures:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{pics/{loops-to-loops-fig5.2}.PNG}
        \caption{Loops to loops (pp. 89 of \cite{book})}
    \end{figure}
    Geometrieally, $\alpha$ winds eaeh of the segments
    $[O,\frac{1}{2}]$, $[\frac{1}{2},\frac{3}{4}]$, $[\frac{3}{4}, 1]$
    once round the eirc1e, the first two being wound in an
    anticlockwise direetion, and the third clockwise. The loop $\beta$
    simply winds the whole interval $[0,1]$ once round the circle
    anticlockwise.

@@

\begin{defi}[Homotopy type]
\nomenclature{Homotopy type}{\nomrefpage.}
    Two spaces $X$ and $Y$ have the same homotopy type, or they are
    homotopy equivalent, if there exist maps:
    $$ \begin{tikzcd}[]
        X\ar[r,"f",bend left] & Y\ar[l,"g",bend left]
    \end{tikzcd}$$
    such that $g\circ f\cong \id_X$, and $f\circ g\cong\id_Y$.
\end{defi}
\begin{fact}
This relationship must be an equivalence relation on topological
spaces, as confirmed by lemma (5.16) on page 103 of \cite{book}.
\end{fact}
\begin{remark}
    The topological spaces $X$ and $Y$ are not necessarily
    path-connected! But we have:
\end{remark}
\begin{fact}
    If $X$ and $Y$ have the same homotopy type then $X$ is
    path-connected if and only if $Y$ is path-connected. (This is
    noted in a footnote on page 107 of \cite{book}.)
\end{fact}
\begin{defi}[Deformation retraction]
\nomenclature{Deformation retraction}{\nomrefpage.}
    Let $A$ be a subspace of $X$. Let a homotopy $G:X\times I\to X$ which
    is relative to $A$ and for which
    $$ \begin{cases}
        G(x,0)=x & \\
        G(x,1)\in A
    \end{cases}$$
    for all $x\in X$. Then $G$ will be called a deformation retraction
    of $X$ onto $A$.
\end{defi}
\begin{remark}
    If there is a deformation retraction of $X$ onto $A$, then of
    course $X$ and $A$ have the same homotopy type.
\end{remark}
Following is an example of a deformation retraction:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{pics/deformation-retraction.PNG}
    \caption{Three Deformation Retractions (from \cite{book})}
\end{figure}

@@

\begin{thm}
    If two \hl{path-connected} spaces are of the same homotopy type, then
    they have isomorphic fundamental group.
\end{thm}
\begin{proof}
    Proof can be found on page 106 of \cite{book}. Here's a note using
    the notation in that book:
    $$ \begin{tikzcd}[]
        X\ar[r,"f",bend left] & Y\ar[l,"g",bend left]
    \end{tikzcd}$$
    It should be noticed that the continuity allows one to identify the
    diagram:
    $$ \begin{tikzcd}[]
        \pi_1(X,p)\ar[r,"(gf)_*"]\ar[rr,"\id",bend right]
            &\pi_1(X,gf(p))\ar[r,"\gamma^{-1}_*"]
            &\pi_1(X,p)
    \end{tikzcd} $$
    with the diagram:
    $$ \begin{tikzcd}[]
        \pi_1(X)\ar[r,"(gf)_*"]\ar[rr,"\id",bend right]
            &\pi_1(X)\ar[r,"\gamma^{-1}_*"]
            &\pi_1(X)
    \end{tikzcd} $$
\end{proof}

\begin{fact}
Using the above fact, we can find that the M\"obius strip, the cylinder,
the punctured plane $\E^2\setminus\{0\}$, and the solid torus, all
have the homotopy type of a circle $S^1$, and consequiently have $\Z$
as fundamental group.
\end{fact}
\begin{fact}
    Also, $\E^n\setminus\{0\}$ deformation-retracts onto $S^{n-1}$ and
    is therefore a simply connected space when $n\geq 3$.
\end{fact}

@@

    \begin{enumerate}
        \item A space is contractible if and only if it has the
            homotopy type of a point.
        \item A contractible space is simply connected.
        \item Any two maps into a contractible space are homotopic.
        \item If $X$ is contractible, then $\id_X$ is homotopic to the
            constant map at $x$ for any $x\in X$.
    \end{enumerate}
\end{thm}
\begin{ex}
    Here is a contractible space that is really hard to imagine. It is
    called the topologist's dunce hat. It is formed by identifying the
    sides of a triangle in the manner indicated in the following
    figure:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.4\linewidth]{pics/topo-dunce-hat.PNG}
        \caption{Topologist's dunce hat}
        \label{fig:topo-d-hat}
    \end{figure}
    You may imagine identifying a pair of sides, then identify the
    identitied edge to the resting side. The direction for identifying
    is really important.
    This space is contractible. Despite some hard words said by the
    book, it is really simple to see on figure~\ref{fig:topo-d-hat}.
    Draw any loop on it and shrink it. Noticing that all edges are
    identified, though in some odd direction.
\end{ex}

\begin{remark}
    An identity function $\id_X$ on a contractible space $X$
    is homotopic to constant function $\id_p$ on a point $p\in X$. But
    this homotopy may not keep the point $p$ fixed. That is, we do not
    necessarily have $\id_X\cong\id_p$~rel $p$.

    An example on page l08 of \cite{book}. In a space called
    the topologist's comb:
    \begin{align*}
        K &= \{\frac{1}{n}|n\in\N\setminus\{0\}\} \\
        \text{comb space}&=(\{0\} \times [0,1] ) \cup (K \times [0,1]) \cup ([0,1] \times \{0\})
    \end{align*}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.6\linewidth]{pics/comb-space.PNG}
        \caption{Topologist's Comb (from Wikipedia)}
    \end{figure}
    We cannot continuously defrom the identity function $\id_X$ to the
    constant function on $p=(0,1)$. Intuitively, there are infinite
    points around $p$ that are infinitesimally close to it. But the
    identity function must shrink from all these infinite points to
    the base line $[0,1]$ without afftecting the value on $p$. This
    sounds unlikely.

    Analytically, the argument is provided by
    \href{http://math.stackexchange.com/users/172604/pedro}{Pedro} in
    his Math.SE
    \href{http://math.stackexchange.com/a/952989/184811}{post}:

    \begin{myquote} \enquote{
        \textbf{Idea:} Take the sequence $x_n=(\frac{1}{n},1)$, it
        converges to $x_0$. If existed such homotopy $H(x,t)$ then the
        sequences $H(x_{n},t)$ would still converge to $x_{0}$.

        You have for each neighborhood $U$ of $x_0$ a number
        $N_{(U,t)}$ and $\epsilon_{(U,t)}$ such that $H(x_n,s)\in U$
        for $n>N_{(U,t)}$ and $|t-s|<\epsilon_{(U,t)}$.

        Covering the interval $I$ with
        $(t-\epsilon_{(U,t)},t+\epsilon_{(U,t)})$ and taking a finite
        subcover you obtain a number $N$ such that $H(x_n,t)\in U$ for
        any $n>N$ and $t\in[0,1]$.

        Now you can use disconnectedness of a small neighborhood to
        show that the homotopy can't take the elements of the sequence
        to $x_0$. 
    } \end{myquote}
\end{remark}
